---
title: "Session 5 Homework Solutions"
subtitle: "APSTA-GE 2006: Applied Statistics for Social Science Research"
author: "Eric Novik"
date: "2026-08-26"
format:
  pdf:
    documentclass: scrartcl
    pdf-engine: lualatex
    papersize: letter
    geometry:
      - margin=0.8in
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    include-in-header:
      text: |
        \providecommand{\E}{}
        \renewcommand{\E}{\mathbb{E}}
        \providecommand{\P}{}
        \renewcommand{\P}{\mathbb{P}}
        \providecommand{\V}{}
        \renewcommand{\V}{\mathbb{V}}
execute:
  echo: true
  warning: false
  message: false
editor: source
---

This document provides worked solutions to the [Lecture 5 homework](https://ericnovik.github.io/stats-math/05-lecture/05-lecture.html#/homework).

```{r}
#| include: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

solution_theme <-
  theme_minimal(base_size = 11) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank()
  )

theme_set(solution_theme)
```

# Binomial PMF and CDF

> **Task:** Let $X\sim\operatorname{Binomial}(20,0.35)$. Compute $\P(X=7)$, $\P(X\leq7)$, and $\P(5\leq X\leq10)$; verify each result with `dbinom()` or `pbinom()`, then plot the PMF and CDF for $x=0,\ldots,20$.

For a binomial random variable with $N=20$ and $\theta=0.35$, the PMF is

$$
\P(X=x)=\binom{20}{x}(0.35)^x(0.65)^{20-x},
\qquad x=0,1,\ldots,20.
$$

Therefore,

$$
\P(X=7)=\binom{20}{7}(0.35)^7(0.65)^{13}.
$$

The two interval probabilities can be written as

$$
\begin{aligned}
\P(X\leq7)
  &=\sum_{x=0}^{7}\binom{20}{x}(0.35)^x(0.65)^{20-x},\\
\P(5\leq X\leq10)
  &=\P(X\leq10)-\P(X\leq4).
\end{aligned}
$$

```{r}
N <- 20
theta <- 0.35

p_equal_7 <- dbinom(7, size = N, prob = theta)
p_at_most_7 <- pbinom(7, size = N, prob = theta)
p_5_through_10 <-
  pbinom(10, size = N, prob = theta) -
  pbinom(4, size = N, prob = theta)

binomial_answers <- data.frame(
  quantity = c("P(X = 7)", "P(X <= 7)", "P(5 <= X <= 10)"),
  probability = c(p_equal_7, p_at_most_7, p_5_through_10)
)

knitr::kable(
  binomial_answers,
  digits = 5,
  booktabs = TRUE,
  caption = "Binomial probability calculations"
)
```

Thus,

$$
\boxed{
\P(X=7)=`r sprintf("%.5f", p_equal_7)`,\quad
\P(X\leq7)=`r sprintf("%.5f", p_at_most_7)`,\quad
\P(5\leq X\leq10)=`r sprintf("%.5f", p_5_through_10)`
}.
$$

The PMF gives each point probability, while the CDF accumulates those probabilities from left to right.

```{r}
#| fig-width: 7.5
#| fig-height: 3.8
#| fig-align: center
#| fig-cap: "PMF and CDF of a Binomial(20, 0.35) random variable."

binomial_grid <- data.frame(
  x = 0:N,
  pmf = dbinom(0:N, size = N, prob = theta),
  cdf = pbinom(0:N, size = N, prob = theta)
)

pmf_plot <- ggplot(binomial_grid, aes(x, pmf)) +
  geom_col(width = 0.7, fill = "#3564d4") +
  scale_x_continuous(breaks = seq(0, N, by = 2)) +
  labs(
    x = "x",
    y = "P(X = x)",
    title = "Probability mass function"
  )

cdf_plot <- ggplot(binomial_grid, aes(x, cdf)) +
  geom_step(direction = "hv", linewidth = 0.6, color = "#d95f02") +
  geom_point(size = 1.3, color = "#d95f02") +
  scale_x_continuous(breaks = seq(0, N, by = 2)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "x",
    y = "P(X <= x)",
    title = "Cumulative distribution function"
  )

pmf_plot + cdf_plot
```

# First-success simulation

> **Task:** Let $T$ be the trial on which the first success occurs when $\theta=0.2$. Simulate 50,000 values with `rgeom() + 1`; for $t=1,\ldots,10$, compare the empirical probabilities with $\P(T=t)=\theta(1-\theta)^{t-1}$. Compare the simulated mean and variance with $1/\theta$ and $(1-\theta)/\theta^2$.

R's `rgeom()` counts the number of failures before the first success. Adding 1 changes that output into the trial number of the first success.

For $T=1,2,\ldots$,

$$
\P(T=t)=\theta(1-\theta)^{t-1},
\qquad
\E(T)=\frac{1}{\theta},
\qquad
\V(T)=\frac{1-\theta}{\theta^2}.
$$

With $\theta=0.2$, the theoretical mean and variance are

$$
\E(T)=5,
\qquad
\V(T)=20.
$$

```{r}
set.seed(502)

geometric_theta <- 0.2
geometric_reps <- 50000
first_success <- rgeom(geometric_reps, prob = geometric_theta) + 1

estimate_empirical_probability <- function(value, simulated_values) {
  number_of_matches <- sum(simulated_values == value)
  number_of_simulations <- length(simulated_values)

  number_of_matches / number_of_simulations
}

geometric_comparison <- data.frame(t = 1:10) |>
  mutate(
    empirical = purrr::map_dbl(
      t,
      estimate_empirical_probability,
      simulated_values = first_success
    ),
    theoretical = geometric_theta * (1 - geometric_theta)^(t - 1),
    difference = empirical - theoretical
  )

knitr::kable(
  geometric_comparison,
  digits = 5,
  booktabs = TRUE,
  caption = "Empirical and theoretical first-success probabilities"
)
```

The pointwise simulation error is small and changes sign across support points, as expected from Monte Carlo variation.

```{r}
#| fig-width: 7
#| fig-height: 4.2
#| fig-align: center
#| fig-cap: "Simulated and theoretical probabilities for the first-success trial."

geometric_comparison |>
  select(t, empirical, theoretical) |>
  pivot_longer(
    cols = c(empirical, theoretical),
    names_to = "source",
    values_to = "probability"
  ) |>
  ggplot(aes(t, probability, color = source, shape = source)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:10) +
  scale_color_manual(
    values = c(empirical = "#3564d4", theoretical = "#d95f02"),
    labels = c(empirical = "simulation", theoretical = "theory")
  ) +
  scale_shape_manual(
    values = c(empirical = 16, theoretical = 1),
    labels = c(empirical = "simulation", theoretical = "theory")
  ) +
  labs(
    x = "trial of first success, t",
    y = "probability",
    color = NULL,
    shape = NULL
  )
```

```{r}
geometric_moments <- data.frame(
  source = c("simulation", "theory"),
  mean = c(mean(first_success), 1 / geometric_theta),
  variance = c(
    var(first_success),
    (1 - geometric_theta) / geometric_theta^2
  )
)

knitr::kable(
  geometric_moments,
  digits = 4,
  booktabs = TRUE,
  caption = "First-success mean and variance"
)
```

The simulated mean is `r sprintf("%.3f", mean(first_success))` and the simulated variance is `r sprintf("%.3f", var(first_success))`, both close to 5 and 20.

# Rare events: binomial and Poisson

> **Task:** Simulate 100,000 values from $\operatorname{Binomial}(1000,0.003)$ and compare the empirical PMF for $0,\ldots,10$ with a Poisson PMF having $\lambda=3$. Compare the two means, variances, and probabilities of at least 6 events.

Here

$$
N=1000,
\qquad
\theta=0.003,
\qquad
N\theta=3.
$$

The binomial mean and variance are

$$
\E(X)=N\theta=3,
\qquad
\V(X)=N\theta(1-\theta)=2.991.
$$

A $\operatorname{Poisson}(3)$ random variable has mean and variance equal to 3, so the two distributions should be close.

```{r}
set.seed(503)

rare_reps <- 100000
rare_size <- 1000
rare_prob <- 0.003
poisson_lambda <- rare_size * rare_prob

rare_counts <- rbinom(
  rare_reps,
  size = rare_size,
  prob = rare_prob
)

# Reuse the empirical-probability function from Problem 2
rare_pmf <- data.frame(k = 0:10) |>
  mutate(
    empirical_binomial = purrr::map_dbl(
      k,
      estimate_empirical_probability,
      simulated_values = rare_counts
    ),
    exact_binomial = dbinom(k, size = rare_size, prob = rare_prob),
    poisson = dpois(k, lambda = poisson_lambda)
  )

knitr::kable(
  rare_pmf,
  digits = 5,
  booktabs = TRUE,
  caption = "Rare-event PMF comparison"
)
```

```{r}
#| fig-width: 7
#| fig-height: 4.3
#| fig-align: center
#| fig-cap: "The empirical binomial PMF and its Poisson comparison."

rare_pmf |>
  pivot_longer(
    cols = -k,
    names_to = "source",
    values_to = "probability"
  ) |>
  ggplot(aes(k, probability, color = source, shape = source)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.8) +
  scale_x_continuous(breaks = 0:10) +
  scale_color_manual(
    values = c(
      empirical_binomial = "#222222",
      exact_binomial = "#3564d4",
      poisson = "#d95f02"
    ),
    labels = c(
      empirical_binomial = "simulated binomial",
      exact_binomial = "exact binomial",
      poisson = "Poisson comparison"
    )
  ) +
  scale_shape_manual(
    values = c(
      empirical_binomial = 16,
      exact_binomial = 1,
      poisson = 2
    ),
    labels = c(
      empirical_binomial = "simulated binomial",
      exact_binomial = "exact binomial",
      poisson = "Poisson comparison"
    )
  ) +
  labs(
    x = "event count",
    y = "probability",
    color = NULL,
    shape = NULL
  )
```

For the upper-tail probability,

$$
\P(X\geq6)=1-\P(X\leq5).
$$

```{r}
rare_summary <- data.frame(
  source = c(
    "simulated binomial",
    "exact binomial",
    "Poisson comparison"
  ),
  mean = c(
    mean(rare_counts),
    rare_size * rare_prob,
    poisson_lambda
  ),
  variance = c(
    var(rare_counts),
    rare_size * rare_prob * (1 - rare_prob),
    poisson_lambda
  ),
  probability_at_least_6 = c(
    mean(rare_counts >= 6),
    pbinom(5, size = rare_size, prob = rare_prob, lower.tail = FALSE),
    ppois(5, lambda = poisson_lambda, lower.tail = FALSE)
  )
)

knitr::kable(
  rare_summary,
  digits = 5,
  booktabs = TRUE,
  caption = "Rare-event summary comparison"
)
```

The exact binomial and Poisson upper-tail probabilities differ by only `r sprintf("%.5f", abs(rare_summary$probability_at_least_6[2] - rare_summary$probability_at_least_6[3]))`. The approximation works well because there are many trials and a small success probability.

# Checking Poisson equidispersion

> **Task:** Repeat 1,000 times: simulate 200 counts from $\operatorname{Poisson}(4)$ and record their sample mean and variance. Plot variance against mean. Then repeat after drawing each count's rate independently from $\{1,7\}$ with equal probability. Compare the plots and explain why the heterogeneous counts are overdispersed relative to a single Poisson model.

For the homogeneous model, every count has the same rate:

$$
X_i\sim\operatorname{Poisson}(4).
$$

Its population mean and variance are both 4. We therefore expect each dataset's sample mean and variance to fluctuate around the diagonal point $(4,4)$.

For the heterogeneous model, each observation first receives rate 1 or rate 7 with equal probability. The average rate remains

$$
\frac{1+7}{2}=4.
$$

However, the rates themselves vary. Their variance is

$$
\frac{(1-4)^2+(7-4)^2}{2}=9.
$$

That between-rate variation adds to the average within-rate Poisson variance of 4, giving an overall count variance of

$$
4+9=13.
$$

Thus, both models have mean 4, but only the homogeneous model has variance 4.

```{r}
set.seed(504)

count_reps <- 1000
counts_per_dataset <- 200

simulate_count_summaries <- function(repetitions, draw_counts) {
  results <- replicate(repetitions, {
    counts <- draw_counts()
    c(
      sample_mean = mean(counts),
      sample_variance = var(counts)
    )
  })

  as.data.frame(t(results))
}

homogeneous <- simulate_count_summaries(
  count_reps,
  function() rpois(counts_per_dataset, lambda = 4)
) |>
  mutate(model = "single rate: lambda = 4")

heterogeneous <- simulate_count_summaries(
  count_reps,
  function() {
    rates <- sample(
      c(1, 7),
      size = counts_per_dataset,
      replace = TRUE
    )
    rpois(counts_per_dataset, lambda = rates)
  }
) |>
  mutate(model = "mixed rates: 1 or 7")

poisson_checks <- bind_rows(homogeneous, heterogeneous)
```

```{r}
poisson_summary <- poisson_checks |>
  group_by(model) |>
  summarise(
    average_sample_mean = mean(sample_mean),
    average_sample_variance = mean(sample_variance),
    sd_of_sample_means = sd(sample_mean),
    .groups = "drop"
  )

poisson_summary_display <- poisson_summary |>
  transmute(
    Model = model,
    `Average mean` = average_sample_mean,
    `Average variance` = average_sample_variance,
    `SD of means` = sd_of_sample_means
  )

knitr::kable(
  poisson_summary_display,
  digits = 3,
  booktabs = TRUE,
  caption = "Repeated-simulation summaries"
)
```

```{r}
#| fig-width: 7.4
#| fig-height: 4.4
#| fig-align: center
#| fig-cap: "Sample variance versus sample mean for homogeneous and heterogeneous Poisson counts."

poisson_checks |>
  ggplot(aes(sample_mean, sample_variance)) +
  geom_abline(
    slope = 1,
    intercept = 0,
    color = "#d95f02",
    linewidth = 0.6,
    linetype = "dashed"
  ) +
  geom_point(color = "#3564d4", alpha = 0.25, size = 0.9) +
  facet_wrap(~ model) +
  scale_x_continuous(
    limits = c(0, 18),
    breaks = seq(0, 18, by = 3),
    expand = expansion(mult = 0.02)
  ) +
  scale_y_continuous(
    limits = c(0, 18),
    breaks = seq(0, 18, by = 3),
    expand = expansion(mult = 0.02)
  ) +
  coord_fixed(ratio = 1) +
  labs(
    x = "sample mean",
    y = "sample variance"
  )
```

Both axes now run from 0 to 18 with a 1:1 aspect ratio, so equal numerical changes occupy equal visual distances. The nearly vertical clouds show that the sample means vary within a narrow range. The homogeneous simulations cluster near mean 4 and variance 4, close to the dashed equidispersion line. The heterogeneous simulations retain mean 4 but cluster near variance 13. A single Poisson model with rate 4 cannot reproduce that extra variation, so the heterogeneous counts are overdispersed relative to it.

# Main conclusions

1. `dbinom()` evaluates a binomial point probability, while `pbinom()` accumulates probabilities through a cutoff.
2. A geometric simulation using `rgeom() + 1` agrees with the first-success PMF, mean, and variance.
3. A Poisson PMF closely approximates a binomial PMF when the number of trials is large, the success probability is small, and $N\theta$ is held fixed.
4. Poisson equidispersion can fail when observations have different rates, even when their average rate is unchanged.
