---
title: "Session 3 Homework Solutions"
subtitle: "APSTA-GE 2006: Applied Statistics for Social Science Research"
author: "Eric Novik"
format:
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - margin=1in
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

This document provides worked solutions to the [Lecture 3 homework](https://ericnovik.github.io/stats-math/03-lecture/03-lecture.html#/homework).

```{r}
#| include: false

library(ggplot2)
library(dplyr)
library(tidyr)

lecture_theme <-
  theme_minimal(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"),
    plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"),
    legend.key = element_rect(fill = "#f0f1eb", color = "#f0f1eb")
  )

theme_set(lecture_theme)
```

# Inspecting the `iris` data

> **Task:** Take a look at the `iris` dataset using `?iris`.

The built-in `iris` data frame contains 150 observations and five variables. Four variables are numeric flower measurements in centimeters, and `Species` identifies one of three iris species. Each species contributes 50 observations.

```{r}
dim(iris)
names(iris)
table(iris$Species)
head(iris)
```

# Overall means

> **Task:** Compute the overall means of `Sepal.Length`, `Sepal.Width`, `Petal.Length`, and `Petal.Width`.

Because the four measurement columns are numeric, `across(where(is.numeric), mean)` applies `mean()` to all four in one step.

```{r}
overall_means <- iris |>
  summarise(across(where(is.numeric), mean))

knitr::kable(
  overall_means,
  digits = 3,
  caption = "Overall means of the iris measurements"
)
```

Thus, the overall means are approximately 5.843 for sepal length, 3.057 for sepal width, 3.758 for petal length, and 1.199 for petal width.

# Means by species

> **Task:** Compute the four means by `Species` using `group_by()` and `summarise()`.

`group_by(Species)` partitions the rows by species. The following `summarise()` call then computes each mean separately within those groups.

```{r}
species_means <- iris |>
  group_by(Species) |>
  summarise(
    across(where(is.numeric), mean),
    .groups = "drop"
  )

knitr::kable(
  species_means,
  digits = 3,
  caption = "Mean iris measurements by species"
)
```

The grouped means reveal substantial differences in petal size: setosa has much shorter and narrower petals than versicolor and virginica.

# Petal-length and sepal-length scatter plot

> **Task:** Reproduce the scatter plot of `Sepal.Length` against `Petal.Length`, colored by `Species`.

Each point represents one flower. Mapping `Species` to color makes the three groups visible, while the axes reproduce the reference plot.

```{r}
#| fig-width: 7
#| fig-height: 5.2
#| fig-align: center
#| fig-cap: "Sepal length versus petal length, colored by species."

scatter_plot <- ggplot(
  iris,
  aes(x = Petal.Length, y = Sepal.Length, color = Species)
) +
  geom_point(size = 2) +
  labs(
    x = "Petal.Length",
    y = "Sepal.Length",
    color = "Species"
  )

scatter_plot
```

The plot shows clear species clustering and a positive relationship between petal length and sepal length.

# Faceted density plots

> **Task:** Reproduce the faceted density plot using `geom_density()` and `facet_wrap()`.

The measurements must first be transformed from wide to long form. In the long data, `name` identifies the measurement and `value` stores its observed value. We then draw one density panel per measurement.

```{r}
#| fig-width: 7
#| fig-height: 5.3
#| fig-align: center
#| fig-cap: "Distributions of the four iris measurements by species."

iris_long <- iris |>
  pivot_longer(
    cols = -Species,
    names_to = "name",
    values_to = "value"
  )

density_plot <- ggplot(
  iris_long,
  aes(x = value, color = Species, fill = Species)
) +
  geom_density(alpha = 1/5) +
  facet_wrap(~ name, scales = "free", ncol = 2) +
  labs(x = NULL, y = "density", color = "Species", fill = "Species")

density_plot
```

Because `name` is a character column, `facet_wrap()` orders the labels alphabetically by default, which matches the reference plot. The free scales are important because petal width and sepal length occupy very different numerical ranges.

# Indefinite integral

> **Task:** Compute the following integral and show the steps:
>
> $$
> \int 2x\cos(x^2)\,dx.
> $$

Use substitution. Let

$$
u=x^2.
$$

Then

$$
du=2x\,dx,
$$

so the integral becomes

$$
\begin{aligned}
\int 2x\cos(x^2)\,dx
  &=\int\cos(u)\,du\\
  &=\sin(u)+C\\
  &=\boxed{\sin(x^2)+C}.
\end{aligned}
$$

Differentiating the result checks the calculation:

$$
\frac{d}{dx}\sin(x^2)=\cos(x^2)(2x)=2x\cos(x^2).
$$

# Definite integral from $0$ to $2\pi$

> **Task:** Evaluate the integral on paper from $0$ to $2\pi$, then validate the result with R's `integrate()` function.

Using the antiderivative above,

$$
\begin{aligned}
\int_0^{2\pi}2x\cos(x^2)\,dx
  &=\left[\sin(x^2)\right]_0^{2\pi}\\
  &=\sin\!\left((2\pi)^2\right)-\sin(0)\\
  &=\boxed{\sin(4\pi^2)}\\
  &\approx0.9783406.
\end{aligned}
$$

Notice that $\sin(4\pi^2)$ is not zero: the argument is $4\pi^2$, not an integer multiple of $2\pi$.

```{r}
integrand <- function(x) {
  2 * x * cos(x^2)
}

numeric_result <- integrate(
  integrand,
  lower = 0,
  upper = 2 * pi
)

numeric_result

c(
  analytic = sin(4 * pi^2),
  numerical = numeric_result$value,
  difference = abs(sin(4 * pi^2) - numeric_result$value)
)
```

The analytical and numerical values agree to numerical precision.
