---
title: "Getting Started with Stan"
layout: post
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path='{{ site.url }}/images/getting-started-')
library(ggplot2)
```

During the recent Bayesian Inference with Stan class, [Daniel Lee](https://www.linkedin.com/in/syclik), one of the Stan's core developers walked us through a series of exercises to help us get started with Stan. This post is my reproduction of his introduction to Stan.

## Installing Stan
If you do not have Stan installed and you are an R User, follow the [RStan Getting Started Instruction](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started). You need to make sure you have a working C++ compiler installed; the directions are provided via the above link. For Mac OS X, the recommendation is to use **clang**. [Interfaces](http://mc-stan.org/interfaces/) also exist for Python, Matlab, Julia, and Stata and for those who prefer to work with the Stan compiler directly, there is Command Stan. In this post, I will only be discussing the R interface.

## Running Stan
Stan program is of course just a text file that you feed to the Stan compiler which translates it into C++ code and compiles it on the native platform. One implication of this is that the compiled Stan object is platform dependent (as opposed to say a Java program that executes on the JVM). 

Once, RStan is installed the package is loaded the usual way:
```{r}
library(rstan)

# enable multicore support so chains can be fit in parallel
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r, cache=TRUE}
data <- list(N = 1000, y = rbinom(1000, 1, 0.7))
sum(data$y) / 1000

bern <- stan("../models/bernoulli.stan", data = data)
bern
pairs(bern)
```

```{r, eval=FALSE}
data {
  int<lower = 0> N;
  int<lower = 0, upper = 1> y[N];
}
parameters {
  real<lower = 0, upper = 1> theta;
}
model {
  theta ~ normal(0.5, 0.2);
  y ~ bernoulli(theta);  
}
```



```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
